---
title: "Rothkopf's Morse Data"
author: "Jan de Leeuw"
date: last-modified
date-format: long
bibliography: [mypubs.bib, total.bib]
number-sections: true
pdf-engine: lualatex
keep-tex: true
editor: source
format:
   pdf:
    fontsize: 12pt
    keep-tex: true
    link-citations: true
    documentclass: scrartcl
    number-sections: true
---

Here are some analyses of the Rothkopf Morse data, using a variety
of loss functions. All analyses are metric and two-dimensional.
We used the assymetric confusion matrix and defined
a dissimilarity by the Shepard-Luce like formula
$$
\delta_{ij}=-\log\frac{p_{ij}p_{ji}}{p_{ii}p_{jj}}.
$$
```{r data, echo = FALSE}
data(morse2, package = "smacof")
morse2 <- 1 - morse2
morse2 <- ifelse(morse2 == 0, .01, morse2)
dm <- diag(morse2)
morse <- -log((morse2 * t(morse2)) / (outer(dm, dm)))
```

The fivenum for these data is
```{r fivenum, echo = FALSE}
fivenum(morse)
```
Thus data are fairly symmetric, a bit skewed to the left, with a
median around four and values between zero and ten,

There are four analyses. 

1. Huber with cut-off 25, well outside the range, de facto least
squares loss.
2. Huber with cut-off .01, de facto least absolute value loss.
3. Huber with cut-off 4, around the median. 
4. Tukey with cut-off 4, around the median.

In the last three analyses, especially in the fourth, the large
residuals are down-weighted. An inifinite number of other analyses
are possible. I am taking requests.

```{r ls, echo = FALSE}
source("smacofRobust.R")
hls <- smacofRobust(morse, engine = smacofHuber, cons = 25, verbose = FALSE, itmax = 10000)
```
```{r pxls, echo = FALSE, fig.align = "center", fig.asp = 2, fig.cap = "Configuration Least Squares"}
par(pty = "s")
plot(hls$x, type = "n", xlab = "dim1", ylab = "dim2")
text(hls$x, row.names(morse))
```

```{r pdls, echo = FALSE, fig.align = "center", fig.asp = 2, fig.cap = "Shepard Plot Least Squares"}
par(pty = "s")
plot(morse, hls$d, xlab = "delta", ylab = "d", col = "RED")
abline(0, 1, col = "BLUE")
```

```{r av, echo = FALSE}
hav <- smacofRobust(morse, engine = smacofHuber, cons = .01, verbose = FALSE, itmax = 10000)
```

```{r pxav, echo = FALSE, fig.align = "center", fig.asp = 2, fig.cap = "Configuration Least Absolute Value"}
par(pty = "s")
plot(hav$x, type = "n", xlab = "dim1", ylab = "dim2")
text(hav$x, row.names(morse))
```

```{r pdav, echo = FALSE, fig.align = "center", fig.asp = 2, fig.cap = "Shepard Plot Least Absolute Value"}
par(pty = "s")
plot(morse, hav$d, xlab = "delta", ylab = "d", col = "RED")
abline(0, 1, col = "BLUE")
```

```{r hmed, echo = FALSE}
hmed <- smacofRobust(morse, engine = smacofHuber, cons = 4, verbose = FALSE, itmax = 10000)
```

```{r pxhmed, echo = FALSE, fig.align = "center", fig.asp = 2, fig.cap = "Configuration Median Huber"}
par(pty = "s")
plot(hmed$x, type = "n", xlab = "dim1", ylab = "dim2")
text(hmed$x, row.names(morse))
```

```{r pdhmed, echo = FALSE, fig.align = "center", fig.asp = 2, fig.cap = "Shepard Plot Median Huber"}
par(pty = "s")
plot(morse, hmed$d, xlab = "delta", ylab = "d", col = "RED")
abline(0, 1, col = "BLUE")
```

```{r htu, echo = FALSE}
htu <- smacofRobust(morse, engine = smacofTukey, cons = 4, verbose = FALSE, itmax = 10000)
```

```{r pxtmed, echo = FALSE, fig.align = "center", fig.asp = 2, fig.cap = "Configuration Median Tukey"}
par(pty = "s")
plot(htu$x, type = "n", xlab = "dim1", ylab = "dim2")
text(htu$x, row.names(morse))
```

```{r pdtmed, echo = FALSE, fig.align = "center", fig.asp = 2, fig.cap = "Shepard Plot Median Tukey"}
par(pty = "s")
plot(morse, htu$d, xlab = "delta", ylab = "d", col = "RED")
abline(0, 1, col = "BLUE")
```