---
title: |
    | Smacof at 50: A Manual
    | Part 5: Unfolding in Smacof
author: 
- Jan de Leeuw - University of California Los Angeles
date: '`r paste("Started December 12 2022, Version of",format(Sys.Date(),"%B %d, %Y"))`'
output:
  bookdown::pdf_document2:
    latex_engine: lualatex 
    includes:
      in_header: preamble.tex
    keep_tex: yes
    toc: true
    toc_depth: 3
    number_sections: yes
  bookdown::html_document2:
    keep_md: yes
    css: preamble.css
    toc: true
    toc_depth: 3
    number_sections: yes
graphics: yes
mainfont: Times New Roman
fontsize: 12pt
bibliography: ["mypubs.bib","total.bib"]
abstract: TBD
---
```{r loadpackages, echo = FALSE}
#suppressPackageStartupMessages (library (foo, quietly = TRUE))
```

```{r load code, echo = FALSE}
#dyn.load("foo.so")
#source("janUtil.R")
```


**Note:** This is a working manuscript which will be expanded/updated
frequently. All suggestions for improvement are welcome. All Rmd, tex,
html, pdf, R, and C files are in the public domain. Attribution will be
appreciated, but is not required. The files can be found at
<https://github.com/deleeuw> in the repositories smacofCode, smacofManual,
and smacofExamples.


## Metric Unfolding

Metric unfolding is a special case of metric MDS, as implemented for example
in smacofAC. Thus we minimize stress defined as
\begin{equation}
\sigma(X)=\sum_{i=1}^{N}\sum_{j=1}^{N}w_{ij}(\delta_{ij}-d_{ij}(X))^2.
\end{equation}
For unfolding
$$
W=
\kbordermatrix{
&n&m\\
n&0&U\\
m&U'&0
}
$$
In other words there is a set of $n$ row-objects and a set of $m$ column-objects. The withiin-set
dissimilarities are missing, or ignored, and only the between-set dissimilarities are fitted.
If we partition $X$ as
$$
X=
\kbordermatrix{
\mbox{\ }&p\\
n&Y\\
m&Z
}
$$
then we can also write
\begin{equation}
\sigma(Y,Z)=\sum_{i=1}^n\sum_{j=1}^mu_{ij}(\delta_{ij}-d(y_i,z_j))^2.
\end{equation}


By defining 
$$
X=
\kbordermatrix{
\mbox{\ }&p\\
n&Y\\
m&Z
}
$$
and
$$
W=
\kbordermatrix{
&n&m\\
n&0&U\\
m&U'&0
}
$$
we can write



The multidimensional unfolding model for preference judgments is often attributed to @coombs, @kruskal_carroll, @roskam



# Initial Configuration

$$
\sigma(C)=\sigma(\tilde C+(C-\tilde C))=\sum_{i=1}^n\sum_{j=1}^m((\delta_{ij}^2-\text{tr}\ A_{ij}\tilde C)-\text{tr}\ A_{ij}(C-\tilde C))^2
$$
$$
\sigma(C)=\sigma(\tilde C)-2\sum_{i=1}^n\sum_{j=1}^m(\delta_{ij}^2-\text{tr}\ A_{ij}\tilde C)\text{tr}\ A_{ij}(C-\tilde C)+\sum_{i=1}^n\sum_{j=1}^m\{\text{tr}\ A_{ij}(C-\tilde C)\}^2
$$

From @deleeuw_groenen_pietersz_U_06
$$
\sum_{i=1}^n\sum_{j=1}^m\{\text{tr}\ A_{ij}(C-\tilde C)\}^2\leq (n+m+2)\text{tr}\ (C-\tilde C)^2
$$
Define 
$$
B(\tilde C):=\frac{1}{n+m+2}\sum_{i=1}^n\sum_{j=1}^m(\delta_{ij}^2-\text{tr}\ A_{ij}\tilde C)A_{ij}
$$
So we minimize
$$
-2\ \text{tr}\ B(\tilde C)C+\text{tr}\ C^2-2\text{tr}\ C\tilde C=\\
\text{tr}\ (C-\{\tilde C + B(\tilde C)\})^2
$$
# Constraints

## Centroid Restriction

$$
\begin{bmatrix}
X\\Y
\end{bmatrix}=
\begin{bmatrix}
I\\
D^{-1}G'
\end{bmatrix}
X
$$


## Linearly Restricted Unfolding

$$
\begin{bmatrix}
X\\Y
\end{bmatrix}=
\begin{bmatrix}
R&0\\
0&S
\end{bmatrix}
\begin{bmatrix}
P\\
Q
\end{bmatrix}
$$
## Normalization Restrictions

$\text{tr}\ X'V_{11}X=1$ or $X'V_{11}X=I$.


# Examples 

## Roskam

## Breakfast

## Gold


# References
