---
title: |
    | Smacof at 50: A Manual
    | Part 8: Homogeneity Analysis with Smacof
author: 
- Jan de Leeuw - University of California Los Angeles
date: '`r paste("Started April 13 2024, Version of",format(Sys.Date(),"%B %d, %Y"))`'
output:
  bookdown::pdf_document2:
    latex_engine: lualatex 
    includes:
      in_header: preamble.tex
    keep_tex: yes
    toc: true
    toc_depth: 3
    number_sections: yes
  bookdown::html_document2:
    keep_md: yes
    css: preamble.css
    toc: true
    toc_depth: 3
    number_sections: yes
graphics: yes
mainfont: Times New Roman
fontsize: 12pt
bibliography: ["mypubs.bib","total.bib"]
abstract: smacofVO
---
```{r loadpackages, echo = FALSE}
#suppressPackageStartupMessages (library (foo, quietly = TRUE))
```

```{r load code, echo = FALSE}
#dyn.load("foo.so")
#source("janUtil.R")
```


**Note:** This is a working manuscript which will be expanded/updated
frequently. All suggestions for improvement are welcome. All Rmd, tex,
html, pdf, R, and C files are in the public domain. Attribution will be
appreciated, but is not required. The files can be found at
<https://github.com/deleeuw> in the repositories smacofCode, smacofManual,
and smacofExamples.


# Simultaneous Non-Metric Unfolding

If we have data where the same objects give preference
judgments over multiple domains, or over the same domain on multiple occasions, or over the same domain under different experimental conditions,
then we can use the stress loss function
\begin{equation}
\sigma(X,Y_1,\cdots,Y_m):=\sum_{j=1}^m\sum_{i=1}^n\min_{\hat d_i^j\in\Delta_i^j}\sum_{l=1}^{k_j}w_{il}^j(\hat d_{il}^j-d(x_i,y_l^j))^2
(\#eq:snmu)
\end{equation}
Note that for each object and variable there are different sets of transformations $\Delta_j$
and for each variable there different matrices of column scores $Y_j$, but there is only a single matrix of row scores $X$.

If the $\Delta_i^j$ contain zero, then the unconstrained minimum of \@ref(eq:snmu)
is clearly zero. Collapsing all $x_i$ and all $y_l^j$ into a single point makes all distances zero, and thus makes stress zero. Some sort of normalization of either $X$ and/or the $Y_j$ is needed to prevent this trivial solution.

There is a extensive literature in the case of non-metric unfolding on preveting trivial solutions. Initially  @roskam_68 and @kruskal_carroll_69 tried to use implicit normalization of stress, which in our context means that
each component of \@ref(eq:snmu)
is given a numerator
\begin{equation}
\sigma(X,Y_1,\cdots,Y_m):=\sum_{j=1}^m\sum_{i=1}^n\frac{\min_{\hat d_i^j\in\Delta_i^j}\sum_{l=1}^{k_j}w_{il}^j(\hat d_{il}^j-d(x_i,y_l^j))^2}{\min_{\hat d_{ij}}\sum_{l=1}^{k_j}w_{il}^j(\hat d_{ij}-d(x_i,y_l^j))^2}
(\#eq:snmuin)
\end{equation}
This is supposed to prevent solutions with 
all $d(x_i,y_l^j)$ are equal within a row of 
the 


In fact, it is easy to see that a minimum of zero is also possible in the situation where the $\Delta_i^j$ have the
subset of all constant vectors (or all non-negative constant vectors). Collapse all
$x_i$ into a single point, and place all $y_l^j$ on a sphere around this point.
Or collapse all $y_l^j$ and put the $x_i$ on a sphere. This makes all $d(x_i,y_l^j)$
equal and thus makes stress zero.

## Homogeneity Analysis

The Gifi System (@gifi_B_90, @michailidis_deleeuw_A_98, @deleeuw_mair_A_09a) presents non-linear or non-metric versions of the classical linear multivariate analysis techniques (regression, analysis of variance, canonical analysis, discriminant analysis, principal component analysis) as special cases of Homogeneity Analysis, also known as Multiple Correspondence Analysis.

We give a somewhat non-standard introduction to homogeneity analysis here, to highlight the
similarities with unfolding and the techniques we will present later on in this paper.

The data are a number of indicator matrices $G_1,\cdots,G_s$. Indicator matrices are binary matrices, with rows that add up to one. They are
used to code categorical variables. Rows corresponds with objects
(or objects), column with the categories (or levels) of a variable.
An element $g_{ij}$ is one in row if object $i$ is in category $j$,
and all other elements in row $i$ are zero.

Homogeneity analysis makes a joint maps in $p$ dimensions of objects
and categories (both represented as points) in such a way that category points are close to the points for the objects in the category. And, vice versa, objects are close to the category points that they score in.
If there is only one variable then it is trivial to make such a 
homogeneous map. We just make sure the object points coincide with
their category points. But there are $s>1$ indicator matrices, corresponding with $s$ categorical variables, and the solution is a compromise trying to achieve homogeneity as well as possible for all variables simultaneously.

Let us use loss function \@ref(eq:snmu) to captures loss of homogeneity in
the sense discussed above.
The sets $\Delta_i^r$ are defined in such a way that $\hat d_{il}^j$ is zero if $i$ is in category $l$ of 
variable $j$. There are no constraints on the other $\hat d$'s in row $i$
of variable $j$. Thus for zero loss we want an object to coincide with all $m$ categories it is in. Under this definition of the $\Delta_i^r$ we have

$$
\min_{\hat d_i^j\in\Delta_i^j}\sum_{l=1}^{k_j}w_{il}^j(\hat d_{il}^j-d(x_i,y_l^j))^2=w_{il(i,j)}^jd(x_i,y_{l(i,j)}^j)^2
$$
where the $l(i,j)$ on the right is the index of the category of variable $j$ that object $i$ is in. In 
homogeneity analysis we actually assume that
for each $i$ and $j$ $w_{il}^j$ is the same for all 
$l$, so that we can simply use $w_{ij}$.


Using indicators we can write loss function ... as
$$
\sigma(X,Y_1,\cdots,Y_m)=
\sum_{j=1}^m\text{tr}\ (X-G_jY_j)'W_j(X-G_jY_j),
$$
In homogeneoty analysis we minimize .. using the
explicit normalization $X'W_\star X=I$. The solution
is
\begin{align}
X\Lambda&=W_\star^{-1}\sum_{j=1}^m W_jG_jY_j,\\
Y_j&=(G_j'W_jG_j)^{-1}G_j'W_jX
\end{align}
where $\Lambda$ is a symmetric matrix of Lagrange
multipliers.


# Loss Function

## The Unconstrained Case

Now consider the closely related problem in which we do not require,
as in homogeneity analysis, that 
$$
\hat d^j_{il(i,j)}=0
$$
but we impose the weaker condition that $\hat d^j_{il(i,j)}$ is less than or equal to all $\hat d^j_{il}$ in row $i$. 


In homogeneity analysis the geometric interpretation of loss is that we
want objects to coincide with the categories they score in (for all variables). The geometric interpretation of loss function ... is that we want 
objects to be closer to the categories they score in than to the categories
they do not score in (for all variables). The plot of the the $k_j$ categories of
variable $r$ defines $k_j$ Voronoi regions. The Voronoi region of
category $\ell$ is the polyhedral convex set of all points closer to category
$\ell$ than to any other category of variable $r$. The loss function
... vanishes if the $x_i$ are all in the Voronoi regions of the categories
they score in (for all variables). It is sufficient for contiguity that the
stars in the star plot are disjoint ? Also that the convex hulls of the
object points are disjoint ?

Minimizing ... over the rows $\delta_i^r$ is a monotone regression for a simple tree order. This is easily handled by using Kruskal's primary approach
to ties (@kruskal_64a, @kruskal_64b, @deleeuw_A_77).

In stage 2 we do one or more metric smacof iterations for given $\Delta_j$
to decrease the loss. These smacof iterations, or Guttman transforms, more or less ignore the fact that we are dealing with a rectangular matrix and use the weights to transform the problem into a symmetric one (as in @heiser_deleeuw_A_79).

Thus for stage two purposes the loss function is
$$
\sigma(Z_1,\cdots,Z_m)=\sum_{j=1}^m\sum_{i=1}^{N_j}\sum_{j=1}^{N_j}w_{ij}^r(\delta_{ij}^r-d_{ij}(Z_j))^2,
$$
with $N_j:=n+k_j$ and

$$
Z_j:=\kbordermatrix{
\mbox{\ }&p\\
n&X\\
k_j&Y_j}.
$$

The weights in $W_j$ are now zero for the two diagonal blocks.


$$
g_i^r:=\kbordermatrix{\mbox{\ }&\ \\n&e_i\\k_j&0}.
$$
$$
h_\ell^r:=\kbordermatrix{\mbox{\ }&\ \\n&0\\k_j&e_\ell}
$$
so that $x_i=Z_j'g_i^r$ and $y_\ell^r=Z_j'h_\ell^r$. It follows that
$$
d^2(x_i,y_\ell^r)=\text{tr}\ Z_j'A^r_{i\ell}Z_j^{\ }
$$
where $A_{i\ell}^r$, of order $n+k_j$, is the rank-one, positive semi-definite, doubly centered matrix 
$$
A_{i\ell}^r:=(g_i^r-h_\ell^r)(g_i^r-h_\ell^r)'
$$
It further follows that
$$
\sum_{i=1}^n\sum_{\ell=1}^{k_j}w_{il}^rd^2(x_i,y_\ell^r)=\text{tr}\ Z_j'V_jZ_j,
$$
where 
$$
V_j:=\sum_{i=1}^n\sum_{\ell=1}^{k_j}w_{il}^rA^r_{i\ell},
$$
so that $V_j$ has the shape
$$
V_j=\kbordermatrix{&n&k_j\\
n&\ddots&\Box\\
k_j&\Box&\ddots}
$$
with $-W_j$ in the off-diagonal block, while the two diagonal
blocks are diagonal matrices with the row and column sums of $W_j$.

In a similar way we define $B_j(Z_j)$, with the same shape as $V_j$, and with
$$
B(Z_j)=\sum_{i=1}^n\sum_{\ell=1}^{k_j}w_{i\ell}^r\frac{\delta_{i\ell}^r }{d_{i\ell}(Z_j)}A^r_{i\ell}
$$

$$
\sigma(Z_1,\cdots,Z_m)=\sum_{j=1}^m\{\eta_j^2-2\text{tr}\ Z_j'B(Z_j)Z_j+Z_j'V_jZ_j\}
$$
with
$$
\eta_j^2:=\sum_{i=1}^{N_j}\sum_{j=1}^{N_j}w_{ij}^r\{\delta_{ij}^r\}^2
$$
Define the Guttman transforms of $Z_j$ as
$$
\tilde Z_j=V_j^+B(Z_j)Z_j
$$
Then
$$
\sigma_j(Z):=\eta_j^2-2\text{tr}\ Z_j'B(Z_j)Z_j+Z_j'V_jZ_j=\eta_j^2-\eta^2(\tilde Z_j)+\text{tr}\ (Z_j-\tilde Z_j)'V_j(Z_j-\tilde Z_j)
$$
If $Z_j^{(k)}$ is $Z_j$ at iteration $k$ then
$$
\text{tr}\ Z_j'B(Z_j)Z_j\geq\text{tr}\ Z_j'B(Z_j^{(k)})Z_j^{(k)}=\text{tr}\ Z_j'V_j\tilde Z_j^{(k)}.
$$
Thus
$$
\sigma_j(Z)\leq\eta_j^2-\eta^2(\tilde Z_j^{(k)})+\text{tr}\ (Z_j-\tilde Z_j^{(k)})'V_j(Z_j-\tilde Z_j^{(k)})
$$
It follows that in iteration $k+1$ we must minimize
$$
\omega(Z_1,\cdots,Z_s):=\sum_{j=1}^m\text{tr}\ (Z_j-\tilde Z_j^{(k)})'V_j(Z_j-\tilde Z_j^{(k)})
$$
to compute the $Z_j^{(k+1)}$.

Computation of the Guttman transform requires Moore-Penrose inverses of the
matrices $V_j$, which are of order $n+k_j$ and extremely sparse. They can also be uncomfortably large because $n$ can be large. It is much more memory-friendly to solve the partioned system
$$
\begin{bmatrix}
R_W&-W\\
-W'&C_W
\end{bmatrix}
\begin{bmatrix}X\\Y\end{bmatrix}=\begin{bmatrix}
R_B&-B\\
-B'&C_B
\end{bmatrix}
$$

More explicitly we must minimize

\begin{multline}
\sum_{j=1}^m\text{tr}\ (X-\tilde X_j)'R_j(X-\tilde X_j)-2\sum_{j=1}^m\text{tr}\ (X-\tilde X_j)'W_j(Y_j-\tilde Y_j)+\\
\sum_{j=1}^m\text{tr}\ (Y_j-\tilde Y_j)'C_j(Y_j-\tilde Y_j)
\end{multline}
where $\tilde X_j$ and $\tilde Y_j$ are the two components of the Guttman
transform $\tilde Z_j$ of the current $Z_j$.

First minimize over $Y_j$ for given $X$. This gives
$$
Y_j=\tilde Y_j-\{V_{22}^j\}^{-1}V_{21}^j(X-\tilde X_j)
$$
$$
X = \{V_{11}^\star\}^{-1}\sum_{j=1}^m\left\{V_{11}^j\tilde X_j-V_{12}^j(Y_j-\tilde Y_j)\right\}
$$

$$
\sum_{j=1}^m V_{11}^j(X-\tilde X_j)+\sum_{j=1}^mV_{12}^j(Y_j-\tilde Y_j)= 0
$$

Or we substitute ... into ... we find that we have to minimize

$$
\sum_{j=1}^m\text{tr}\ (X-\tilde X_j)'(R_j-W_jC_j^{-1}W_j')(X-\tilde X_j)
$$
over $X$.
## Missing Data


## Normalization of X

Constraint either (weak)
$$
\text{tr}\ X'V_{11}^\star X = 1
$$
or (strong)
$$
X'V_{11}^\star X = I
$$

## The Unweighted Case

In the  unweighted case we have $V_{11}^r=k_jI$, $V_{22}^r=nI$, and
$V_{12}^r=-E_{n\times k_j}$. Thus
$V_{1\mid2}^r=k_jJ_n$ and $V_{1\mid 2}^\star=k_\star J_n$
with $J_n=I_n-n^{-1}E_{n\times n}$, the centering matrix of order $n.

The Guttman transform also simplifies in the unweighted case. The formulas were already given in @heiser_deleeuw_A_79.

## Centroid Constraints on Y

$$
Z_j=
\kbordermatrix{
\mbox{\ }&p\\
n&I\\
k_j&D_j^{-1}G_j'}X=H_jX
$$
$$
\sum_{j=1}^m Z_j'V_jZ_j=X'\{\sum_{j=1}^mH_j'V_jH_j\}X.
$$

$$
X'H_j'V_jH_jX=X'\{V_{11}^j+V_{12}^jD_j^{-1}G_j+G_jD_j^{-1}V_{22}^jD_j^{-1}G_j'\}X
$$

## Rank Constraints on Y

$$
y_j^r=\alpha_{j}^ry_j
$$

# Note

In order to not have to deal with general inverses and multiplications of gigantic symmetric matrices with largely empty diagonal blocks we solve the system
$$
\begin{bmatrix}
W_j&-W\\
-W'&W_c
\end{bmatrix}
\begin{bmatrix}
X\\Y
\end{bmatrix}
=
\begin{bmatrix}
P\\Q
\end{bmatrix}
$$
for $X$ and $Y$, with the known right hand side
$$
\begin{bmatrix}P\\Q\end{bmatrix}=
\begin{bmatrix}
B_j&-B\\
-B'&B_c
\end{bmatrix}
\begin{bmatrix}
X^{(k)}\\Y^{(k)}
\end{bmatrix}
$$

$W_jX-WY=P$ or $X=W_j^{-1}(P+WY)$. Substitute in $W_cY-W'X=Q$
to get $W_cY-W'W_j^{-1}(P+WY)=Q$ or $(W_c-W'W_j^{-1}W)Y=Q+W'W_j^{-1}P$.

Note that $W_c-W'W_j^{-1}W$ is doubly-centered and $Q+W'W_j^{-1}P$ is column-centered.

# References
