---
title: |
    | Smacof at 50: A Manual
    | Part x: Linearly Constrained Smacof
author: 
- Jan de Leeuw - University of California Los Angeles
date: '`r paste("Started May 08, 2024, Version of",format(Sys.Date(),"%B %d, %Y"))`'
output:
  bookdown::pdf_document2:
    latex_engine: lualatex 
    includes:
      in_header: preamble.tex
    keep_tex: yes
    toc: true
    toc_depth: 3
    number_sections: yes
  bookdown::html_document2:
    keep_md: yes
    css: preamble.css
    toc: true
    toc_depth: 3
    number_sections: yes
graphics: yes
mainfont: Times New Roman
fontsize: 12pt
bibliography: ["mypubs.bib","total.bib"]
abstract: TBD
---
```{r loadpackages, echo = FALSE}
#suppressPackageStartupMessages (library (foo, quietly = TRUE))
```

```{r load code, echo = FALSE}
#dyn.load("foo.so")
#source("janUtil.R")
```


**Note:** This is a working manuscript which will be expanded/updated
frequently. All suggestions for improvement are welcome. All Rmd, tex,
html, pdf, R, and C files are in the public domain. Attribution will be
appreciated, but is not required. All files can be found at
<https://github.com/deleeuw> in the repositories smacofCode, smacofManual,
and smacofExamples. 

# Introduction

$$
X=YA=\sum_{j=1}^m Y_jA_j
$$
with $Y_j$  known $n\times k_j$ "design" matrices and $A$
an unknown $k_j\times p$ matrix of coefficients.

Various constraints on $A$.

1. No constraints
2. $A_j$ are diagonal
2. $A_j$ are scalar
3. $A_j$ are block diagonal

Special choices of $Y_j$

1. Previous iterations of $X$
2. Indicator matrices
3. Design matrices (example the rectangles)

In general the constraint phase must solve minimization of
$$
\omega(A):=\text{tr}\ (\overline{X}-\sum_{j=1}^m Y_jA_j)'V(\overline{X}-\sum_{j=1}^m Y_jA_j)
$$
Minimize
$$
-2\sum_{j=1}^m \text{tr}\ A_j'Y_j'V\overline{X}+
\sum_{j=1}^m\sum_{l=1}^m\text{tr}\ A_j'Y_j'VY_l^{\ }A_l^{\ }
$$

diagonal (not very interesting)

$$
-2\sum_{j=1}^m a_j'\text{diag}\{Y_j'V\overline{X}\}+
\sum_{j=1}^m\sum_{l=1}^m a_j'\text{diag}\{Y_j'VY_l^{\ }\}a_l
$$
scalar
$$
-2\sum_{j=1}^m \alpha_j\text{tr}\ Y_j'V\overline{X}+
\sum_{j=1}^m\sum_{l=1}^m\alpha_j\alpha_l\text{tr}\ Y_j'VY_l^{\ }
$$
rank one $A_j=z_jt_j'$ (not very interesting)


$$
-2\sum_{j=1}^mz_j'Y_j'V\overline{X}t_j+
\sum_{j=1}^m\sum_{l=1}^mt_l't_j\ z_j'Y_j'VY_l^{\ }z_l
$$

block diagonal

$$
A=\bigoplus_{j=1}^m A_j
$$

$$
X=(Y_1A_1\cdots Y_mA_m)
$$

$$
\omega(A)=\sum_{j=1}^m\text{tr}\ (\overline X_j-Y_jA_j)'V(\overline X_j-Y_jA_j)
$$


# References
