---
title: |
    | Smacof at 50: A Manual
    | Part x: Linearly Constrained Smacof
author: 
- Jan de Leeuw - University of California Los Angeles
date: '`r paste("Started May 08, 2024, Version of",format(Sys.Date(),"%B %d, %Y"))`'
output:
  bookdown::pdf_document2:
    latex_engine: lualatex 
    includes:
      in_header: preamble.tex
    keep_tex: yes
    toc: true
    toc_depth: 3
    number_sections: yes
  bookdown::html_document2:
    keep_md: yes
    css: preamble.css
    toc: true
    toc_depth: 3
    number_sections: yes
graphics: yes
mainfont: Times New Roman
fontsize: 12pt
bibliography: ["mypubs.bib","total.bib"]
abstract: TBD
---
```{r loadpackages, echo = FALSE}
#suppressPackageStartupMessages (library (foo, quietly = TRUE))
```

```{r load code, echo = FALSE}
#dyn.load("foo.so")
#source("janUtil.R")
```


**Note:** This is a working paper which will be expanded/updated frequently. All suggestions for improvement are welcome. 

$$
X=YA=\sum_{j=1}^m Y_jA_j
$$
with $Y_j$ a known $n\times k_j$ "design" matrix and $A$
an unknown $k_j\times p$ matrix of coefficients.

Various constraints on $A$.

1. No constraints
2. $A_j$ are diagonal
3. $A_j$ are scalar
4. $A_j$ have rank one.

Special choices of $Y_j$

1. Previous iterations of $X$
2. Indicator matrices
3. Design matrices (example the rectangles)

In general the constraint phase must solve minimization of
$$
\omega(A):=\text{tr}\ (\overline{X}-\sum_{j=1}^m Y_jA_j)'V(\overline{X}-\sum_{j=1}^m Y_jA_j)
$$


# References
